---
output:
  html_document: default
  pdf_document: default
---
#Practical Machine Learning - Week 4 Course Project

###Lionel Lawrence, 12th December 2017

##OVERVIEW

To predict the manner in which the participants did the exercise. This is the "classe" variable in the training set. Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. 

##SETUP
```{r}
#install.packages("randomForest")
#install.packages("e1071")
#install.packages("caret")
suppressWarnings(suppressMessages(library(caret)))
suppressWarnings(suppressMessages(library(randomForest)))
suppressWarnings(suppressMessages(library(e1071)))
set.seed(1603)
```


##TASK

Create a report describing

* how you built your model,
* how you used cross validation
* what you think the expected out of sample error is
* why you made the choices you did

#INPUT

###Download source data

```{r}
trainingFilename   <- 'pml-training.csv'
quizFilename       <- 'pml-testing.csv'
trainingUrl        <- 'https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv'
quizUrl            <- 'https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv'

download.file(trainingUrl, trainingFilename)
download.file(quizUrl,quizFilename)
```


#DATA CLEANING


###Cleaning the null and blank values. Removing these from training data...
```{r}
training.df     <-read.csv(trainingFilename, na.strings=c("NA","","#DIV/0!"))
training.df     <-training.df[,colSums(is.na(training.df)) == 0]
dim(training.df)

```

##Test data cleaning

```{r}
quiz.df         <-read.csv(quizFilename , na.strings=c("NA", "", "#DIV/0!"))
quiz.df         <-quiz.df[,colSums(is.na(quiz.df)) == 0]
dim(quiz.df)
```

#FEATURES

###Dimensionality reduction
```{r}
Training.df   <-training.df[,-c(1:7)]
Quiz.df <-quiz.df[,-c(1:7)]
dim(Training.df)
```

#ALGORITHM

##Separating the training data into training/validation set

```{r}
inTrain     <- createDataPartition(Training.df$classe, p = 0.6, list = FALSE)
inTraining  <- Training.df[inTrain,]
inTest      <- Training.df[-inTrain,]
dim(inTraining);dim(inTest)
```
##Model construction using cross validation

```{r}
myModelFilename <- "myModel.RData"
library(doParallel)
ncores <- makeCluster(detectCores() - 1)
registerDoParallel(cores=ncores)
getDoParWorkers() # 3    

# use Random Forest method with Cross Validation, 4 folds
myModel <- train(classe ~ .
        ,data = inTraining
        ,method = "rf"
        ,metric = "Accuracy"  # categorical outcome variable so choose accuracy
        ,preProcess=c("center", "scale") # attempt to improve accuracy by normalising
        ,trControl=trainControl(method = "cv"
                                ,number = 4 # folds of the training data
                                ,p= 0.60
                                ,allowParallel = TRUE 
                                )
        )

save(myModel, file = myModelFilename)

stopCluster(ncores)
```

```{r}
print(myModel, digits=4)
```

#PREDICT

###Predicting the activity performed using test file derived from training set

```{r}
predTest <- predict(myModel, inTest)
```

#EVALUATIOn

##Test
###Checking the accuracy of the model by comparing the predictions to the actual results

```{r}
confusionMatrix(predTest, as.factor(inTest$classe))
```

#Out of Sample Error
The out-of-sample error is 0.0037 or 0.37%.

Accuracy is very high at 0.9963 and this figure lies within the 95% confidence interval.

#Final Model data and important predictors in the model

```{r}
myModel$finalModel
```
```{r}
varImp(myModel)
```

#Validation/Quiz
The accuracy of the model by predicting with the Validation/Quiz set supplied in the test file

```{r}
print(predict(myModel, newdata=Quiz.df))
```

